On the propriety of restricted Boltzmann machines

### 2016-10-31 writing/draft.Rmd
update some things after oral

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_H, v_1,\dots,v_V)$ represent the states of the visible and hidden nodes in an RBM for some integers $V, H \ge 1$. Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural or na{\"i}ve starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_H, V_1,\dots,V_V)$, for the nodes as

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{X} 
\end{align}

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{H+V}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For 
\begin{align}
\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_H, v_1, \dots, v_V, v_1 h_1, \dots, v_V h_H) \in \mathcal{C}^{H + V + H*V}, \boldsymbol x = (h_1, \dots, h_H, v_1, \dots, v_V) \in \mathcal{C}^{V + H}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$ be the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ for the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, V,\\j = 1, \dots, H}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.

@@

The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows. When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{H + V} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\}.
$$

@@

  return(plot.data)
}
indep_params <- function(samp, H, V) {
  samp[, (H + V + 1):ncol(samp)] <- 0
  samp
}
max_Q <- function(theta, stats) {
  apply(crossprod(t(stats), theta), 2, max)
}
min_Q <- function(theta, stats) {
  apply(crossprod(t(stats), theta), 2, min)
}
min_max_Q <- function(theta, stats) {
  require(dplyr)
  tcrossprod(stats, theta) %>% data.frame() %>%
    cbind(stats) %>%
    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
    summarise_each(funs(max), contains("X")) %>%
    ungroup() %>%
    summarise_each(funs(min), contains("X")) %>%
    select(contains("X")) %>%
    data.matrix() %>%
    t()
}

@@

#instability
res %>%
  group_by(H, V, N, r1, r2) %>%
  do(data.frame(max = max_Q(t(.$samp[[1]]), .$stat[[1]]),
                min = min_Q(t(.$samp[[1]]), .$stat[[1]]),
                min_max = min_max_Q(.$samp[[1]], .$stat[[1]]))) %>%
  ungroup() %>% 
  group_by(H, V, N, r1, r2) %>%
  mutate(LHS1 = (max - min)/N,
         LHS2 = (max - min_max - H*log(2))/N) %>%
  summarise_each(funs(mean), LHS1, LHS2) -> max_q_summary

convex_hull_summary %>%
  left_join(max_q_summary) %>%
  left_join(exp_vals_summary) -> three_ways

@@

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = LHS2)) +
  geom_contour(aes(x = r1, y = r2, z = LHS2), colour = "black", bins = 8) +
  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(frac(bar(R(theta)), V)), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
  theme(aspect.ratio = 1) -> p.min_max_q

@@

```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $R(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size. As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=4}
p.min_max_q
```

>>>>

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$. Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural or na{\"i}ve starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes as

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align}

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For 
\begin{align}
\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\}$ be the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.

@@

The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows. When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\}.
$$

@@

  return(plot.data)
}
indep_params <- function(samp, H, V) {
  samp[, (H + V + 1):ncol(samp)] <- 0
  samp
}
max_Q <- function(theta, stats) {
  apply(crossprod(t(stats), theta), 2, max)
}
min_Q <- function(theta, stats) {
  apply(crossprod(t(stats), theta), 2, min)
}
min_max_Q <- function(theta, stats) {
  require(dplyr)
  tcrossprod(stats, theta) %>% data.frame() %>%
    cbind(stats) %>%
    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
    summarise_each(funs(max), contains("X")) %>%
    ungroup() %>%
    summarise_each(funs(min), contains("X")) %>%
    select(contains("X")) %>%
    data.matrix() %>%
    t()
}
elpr <- function(theta, stats) {
  require(dplyr)
  exp(tcrossprod(stats, theta)) %>% data.frame() %>%
    cbind(stats) %>%
    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
    summarise_each(funs(sum), contains("X")) %>%
    ungroup() -> marginalized

  marginalized %>%
    summarise_each(funs(max), contains("X")) %>%
    select(contains("X")) %>%
    data.matrix() -> max_marg

  marginalized %>%
    summarise_each(funs(min), contains("X")) %>%
    select(contains("X")) %>%
    data.matrix() -> min_marg

  t(log(max_marg/min_marg))
}

@@

#instability
res %>%
  group_by(H, V, N, r1, r2) %>%
  do(data.frame(max = max_Q(t(.$samp[[1]]), .$stat[[1]]),
                min = min_Q(t(.$samp[[1]]), .$stat[[1]]),
                min_max = min_max_Q(.$samp[[1]], .$stat[[1]]))) %>%
  ungroup() %>% 
  group_by(H, V, N, r1, r2) %>%
  mutate(LHS1 = (max - min)/V,
         LHS2 = (max - min_max - H*log(2))/V) %>%
  summarise_each(funs(mean), LHS1, LHS2) -> max_q_summary

res %>%
  group_by(H, V, N, r1, r2) %>%
  do(data.frame(elpr = elpr(.$samp[[1]], .$stat[[1]]))) %>%
  ungroup() %>%
  group_by(H, V, N, r1, r2) %>% 
  summarise(mean_elpr = mean(elpr)) %>%
  mutate(scaled_mean_elpr = mean_elpr/V) -> elpr_summary

convex_hull_summary %>%
  left_join(elpr_summary) %>%
  left_join(exp_vals_summary) -> three_ways

@@

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(frac(bar(ELPR(theta)), V)), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
  theme(aspect.ratio = 1) -> p.elpr

@@

```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $R(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size. As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=4}
p.elpr
```
